---
title: "STATS 604 Project 2 EDA"
author: "Ethan Schubert"
date: "2025-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval=TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(pheatmap)
```

```{r, eval=T}
# Load the data
load("../data/brain.rda")
```

Although the data consists of 386 tissue samples analyzed via microarrays, the samples only come from 86 patients.
```{r, eval=F}
# Number of patients
length(unique(arraymeta$patient))
```

The distribution of observed ages is left-skewed with the majority of ages between 40 and 70 years. The skew is slightly more apparent when comparing the distribution of ages per patient. Furthermore, by comparing the two distributions it seems that more brain tissue samples tended to be collected from younger patients (below 70). Of the 386 observations, only 66 (17.1%) are samples from patients 70 years or older.

```{r, eval=T, fig.show='hold', out.width="49%"}
# Histogram of ages (by observation)
hist(arraymeta$age, main="Distribution of observed ages", xlab="Age (years)")
patients = unique(arraymeta[,c("patient", "age", "sex")])
hist(patients$age, main="Distribution of patient ages", xlab="Age (years)")
```

There are some imbalances in proportions across different metadata variables as well. First, 75.4% of samples are from male patients, while only 24.6% are from female patients. Additionally, array version 1 was used for 62% of observations and 38% were array version 2. The proportion of samples from each region is also imbalanced, with 21% of samples collected from the cerebellum, 32.9% from the anterior cingulate cortex, and 46.1% from the dorsolateral prefrontal cortex. The proportions of samples analyzed at each lab is more balanced, but the University of Michigan analyzed the most at 42.2% while UC Davis and UC Irvine each analyzed about 30%.

```{r, eval=T}
# Relative frequency of each array version
v2_prop = mean(arraymeta$arrayversion-1)

# Relative frequency of each sex (per observation). Much higher proportion of male brains
prop_female = mean(arraymeta$sex=="F")

metadata_factors = c("ANCG", "CB", "DLPFC", "UC Davis", "UC Irvine", "U Michigan", "Female", "Array Version 2")
props = c(colMeans(arraymeta[,2:7]), prop_female, v2_prop)
freq_table = data.frame(factor = metadata_factors, proportion = props)
freq_table %>% kbl() %>%
  kable_paper("hover", full_width = F)
```

More concerning, however, are imbalances of regions across labs. Each lab has low counts for a different brain region. In fact, the Davis lab did not have any samples from the anterior cingulate cortex.

```{r, eval=T}
# Frequencies of region by lab
freq_table = matrix(nrow=3, ncol=3)
rownames(freq_table) = colnames(arraymeta)[2:4]
colnames(freq_table) = colnames(arraymeta)[5:7]
for(i in 2:4) {
  ix = which(arraymeta[,i] == 1)
  freq_table[i-1,] = colSums(arraymeta[ix, 5:7])
}
colnames(freq_table) = c("UC Davis", "UC Irvine", "U Mich")
rownames(freq_table) = c("ANCG", "CB", "DLPFC")
freq_table %>% kbl() %>%
  kable_paper("hover", full_width = F)
```

Additionally, microarrays are designed so that some of the probesets are controls which detect bacterial (non-human) genes. A fixed amount of these bacterial genes are placed into each microarray, and so theoretically the brightness for these genes should be similar. In reality, the mean brightness varies substantially across controls, as the side-by-side boxplots show.

```{r, eval=T}

ii = which(substr(colnames(genemeta),1,4)=="AFFX")
control_means = apply(expression[,ii], 2, mean)
probeset_means = apply(expression[,-ii], 2, mean)

df.means = data.frame(mean = control_means, type = "control")
df.means = rbind(df.means, cbind(mean = probeset_means, type="probeset"))
rownames(df.means) = NULL

df.means %>% ggplot(aes(x = type, group = type, y = mean)) +
  geom_boxplot() +
  ylab("Mean Expression") +
  labs(title="Mean expression by group") +
  theme(axis.text.y=element_blank())
```

We also visualize the expression level for all observations across all 22277 probesets. In the heatmap, each row represents a different observation and each column is a different probeset. The columns are standardized to account for inherent brightness differences between probesets. A clear banding pattern emerges across rows, which suggests that if a sample exhibits low (or high) expression for one probeset, the sample is likely to be consistently low (or high) for other probesets, relative to the other samples. Additionally, the rows are sorted by lab and region. Within some lab and region combinations, the expression levels appear to be either uniformly low or high. This could indicate the possible presence of lab or region effects.

```{r, eval=T}
# Make heatmaps of expression data
lab_region_heatmap = function(data, metadata, scale="none") {
  ix = order(metadata$lab.davis, metadata$lab.irvine, metadata$lab.michigan, 
           metadata$region.ancg, metadata$region.cb, metadata$region.dlpfc)
  labregion.symb = c(rep("M+D", 84), rep("M+C", 18), rep("M+A", 61),
                   rep("I+D", 9), rep("I+C", 42), rep("I+A", 66),
                   rep("D+D", 85), rep("D+C", 21))
  exp_hmap = data[ix,]
  eight_nums = (1:48)*8-7
  rn = rep("", nrow(exp_hmap))
  rn[eight_nums] = labregion.symb[eight_nums]
  rownames(exp_hmap) = rn
  
  colors = hcl.colors(7, palette = "RdYlBu", rev=T)
  p = pheatmap(mat = exp_hmap, color = colors, scale=scale, cluster_rows=FALSE, cluster_cols=FALSE, show_rownames = TRUE, show_colnames=F)
  p
}

lab_region_heatmap(expression, arraymeta, scale="column")
```

